---
title: "Reproducible Research Project 2"
author: "Alan Jenks"
date: "19/05/2019"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Synopsis

Storms and other severve weather conditions not only create a huge econdomic problem but can be extremely harmful for popultation health. We consistently see with a variety of severe weather condtions injuries, fatalities, property & food damage which can be devistating to regions or countries.

The project looks at data from the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm datatbase.

The database contains data from 1950 to the end of November 2011. The data is an accumulation of events and characteristics of storms in the United States looking at where they occur, damage and fatalities and injuries.

This report analyzes the storm event records in the severe weather tracking database to determine which of these event types has the most impact in terms of harm to human health and property damage.

The analysis results show that Tornado events are associated with the vast majority of weather-related human and economic impact. It also shows flood to be the major economic problem for property and drought as the major problem for food.

###Data Processing

Data processing will involve first loading the libraries I will use for this project. I will then download the zip file, unzip it and process the file into a table. I will run some preliminary views of the table to look for items I can pull out and evaluate for the two questions which are which weather condition or conditions has the most impact on personal healh and which storm conditions have the greatest economic impact.

Once the files are downloaded and put in tables. I will isolate the needed variables and clean the data to look at creating some initial plots to see what the dat says.

## Load Libraries


```{r libraries, echo=FALSE}
library(data.table)
library(dplyr)
library(plyr)
library(stats)
library(R.utils)
library(RColorBrewer)
library(knitr)
library(markdown)
library(rmarkdown)
library(tools)
```

## Get Data

#Downaload raw data and read the csv

```{r Data, echo=TRUE}
fileurl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
if (!file.exists('./repdata%2Fdata%2FStormData.csv.bz2')){
  download.file(fileurl,'./repdata%2Fdata%2FStormData.csv.bz2', mode = 'wb')
  bunzip2("repdata%2Fdata%2FStormData.csv.bz2", "repdata%2Fdata%2FStormData.csv", remove = FALSE, skip = TRUE)
}
##read file
storm <- read.csv('repdata%2Fdata%2FStormData.csv')
dim(storm)
```
Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

902297 data points in 37 columns

```{r head, echo=TRUE}
head(storm)
```

```{r names, echo=TRUE}
names(storm)
```

```{r str, echo=TRUE}
str(storm)
```

The dataset has a total of 902297 records

This report focuses on the information related to:
- Type of Event recorded in column 8 labeled EVTYPE
- Number of fatalities recorded in column 23 labeled FATALITIES
- Number of injuries recorded in column 24 labeled INJURIES
- Amount of property damage recorded in column 25 & 26, labeled PROPDMG & PROPDMGEP
- Amount of crop damage recorded in columnS 27 & 28 labeled CROPDMG & CROPDMGEXP

##Code to subset the dataset for fields of interest

```{r subset, echo=TRUE}
sd <- subset(storm[,c(8,23,24,25,26,27,28,29)])
```

###1. Across the United States, which types of events (as indicated in the \color{red}{\verb|EVTYPE|}EVTYPE variable) are most harmful with respect to population health?

##To determine which type of events are most harmful to the population health we must look at the variables fatalities and Injuries.

#Fatalities

```{r fatalities, echo=TRUE}
sumFatalities <- aggregate(FATALITIES ~ EVTYPE, data = sd,  FUN="sum")
dim(sumFatalities)
```


```{r fatalities, echo=TRUE}
fatalitiesTop10 <- sumFatalities[order(-sumFatalities$FATALITIES), ][1:10, ]
dim(fatalitiesTop10)
```


```{r fatalities, echo=TRUE}
fatalitiesTop10
```


```{r fatalities, echo=TRUE}
par(mfrow = c(1,1), mar = c(12, 4, 3, 2), mgp = c(3, 1, 0), cex = 0.8)
barplot(fatalitiesTop10$FATALITIES, names.arg = fatalitiesTop10$EVTYPE, las = 3, main = "10 Fatalities Highest Events", ylab = "Number of Fatalities", col = c(brewer.pal(n = 10, name = "Set2")))

dev.copy(png, "cropdmg-total.png", width = 480, height = 480)
dev.off()
```
#Injuries

```{r injuries, echo=TRUE}
sumInjuries <- aggregate(INJURIES ~ EVTYPE, data = storm,  FUN="sum")
dim(sumInjuries)
```


```{r injuries, echo=TRUE}
injuriesTop10 <- sumInjuries[order(-sumInjuries$INJURIES), ][1:10, ]
dim(injuriesTop10)
```


```{r injuries, echo=TRUE}
injuriesTop10
```


```{r injuries, echo=TRUE}
par(mfrow = c(1,1), mar = c(12, 6, 3, 2), mgp = c(4, 1, 0), cex = 0.8)
barplot(injuriesTop10$INJURIES, names.arg = injuriesTop10$EVTYPE, las = 3, main = "10 Injuries Highest Events", ylab = "Number of Injuries", col = c(brewer.pal(n = 10, name = "Set2")))

dev.copy(png, "cropdmg-total.png", width = 480, height = 480)
dev.off()
```

###2. Across the United States, which types of events have the greatest economic consequences?

##Property Damage

```{r converting damage to dollars, echo=TRUE}
unique(storm$PROPDMGEXP)
storm$PROPDMGEXP <- mapvalues(storm$PROPDMGEXP, from = c("K", "M","", "B", "m", "+", "0", "5", "6", "?", "4", "2", "3", "h", "7", "H", "-", "1", "8"), to = c(10^3, 10^6, 1, 10^9, 10^6, 0,1,10^5, 10^6, 0, 10^4, 10^2, 10^3, 10^2, 10^7, 10^2, 0, 10, 10^8))
storm$PROPDMGEXP <- as.numeric(as.character(storm$PROPDMGEXP))
storm$PROPDMGTOTAL <- (storm$PROPDMG * storm$PROPDMGEXP)/1000000000
unique(storm$CROPDMGEXP)
storm$CROPDMGEXP <- mapvalues(storm$CROPDMGEXP, from = c("","M", "K", "m", "B", "?", "0", "k","2"), to = c(1,10^6, 10^3, 10^6, 10^9, 0, 1, 10^3, 10^2))
storm$CROPDMGEXP <- as.numeric(as.character(storm$CROPDMGEXP))
storm$CROPDMGTOTAL <- (storm$CROPDMG * storm$CROPDMGEXP)/1000000000
```

```{r converting damage to dollars, echo=TRUE}
sumPropertyDamage <- aggregate(PROPDMGTOTAL ~ EVTYPE, data = storm, FUN="sum")
dim(sumPropertyDamage)
```

```{r converting damage to dollars, echo=TRUE}
propdmg10Total <- sumPropertyDamage[order(-sumPropertyDamage$PROPDMGTOTAL), ][1:10, ]
propdmg10Total
```



```{r poperty damage, echo=TRUE}
par(mfrow = c(1,1), mar = c(12, 6, 3, 2), mgp = c(3, 1, 0), cex = 0.8)
barplot(propdmg10Total$PROPDMGTOTAL, names.arg = propdmg10Total$EVTYPE, las = 3, main = "10 Property Damages Highest Events", ylab = "Damage Property Values (in Billions)", col = c(brewer.pal(n = 10, name = "Dark2")))

dev.copy(png, "cropdmg-total.png", width = 480, height = 480)
dev.off()
```

##Crop Damage

```{r crop damage, echo=TRUE}
sumCropDamage <- aggregate(CROPDMGTOTAL ~ EVTYPE, data = storm,  FUN="sum")
dim(sumCropDamage)
```



```{r crop damage, echo=TRUE}
cropdmg10Total <- sumCropDamage[order(-sumCropDamage$CROPDMGTOTAL), ][1:10, ]
cropdmg10Total
```


```{r crop damage, echo=TRUE}
par(mfrow = c(1,1), mar = c(10, 6, 3, 2), mgp = c(3, 1, 0), cex = 0.6)
barplot(cropdmg10Total$CROPDMGTOTAL, names.arg = cropdmg10Total$EVTYPE, las = 2, main = "10 Crop Damages Highest Events", ylab = "Damage Crop Values (in Billions) ", col = c(brewer.pal(n = 10, name = "Dark2")))

dev.copy(png, "cropdmg-total.png", width = 480, height = 480)
dev.off()
```


###Results

## Question 1 - Which is the most harmful with respect to population health?

When we look at injuries and fatalities based on the storm data Tornados are the clear winner as far as the most harmful with respect to population health.

## Question 2 - Across the United States, which types of events have the greatest economic consequences?

Looking at the economic variables which would relate to crop damage and property damage, the data support that floods are the winner with regards to property damage economically and with regards to crop damage drought is by far the winner.
